Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-04-04 vertical-thermal-review.Rmd
metabolism --> gas exchange
we use these pretty interchangeably, but in some cases gas exchange is more suitable.

#Word count
w_intro=369
w_biophys=1515
w_tleaf=749
w_traits= 2391
w_met=940
w_ecol=1353
w_vertgradients=w_biophys+w_tleaf+w_traits+w_met+w_ecol
w_imp_gc=1088
w_imp_modeling_scaling=1186
w_implications=w_imp_gc+w_imp_modeling_scaling
w_conclusions=248
w_total=w_intro+w_vertgradients+w_implications+w_conclusions

@@

![**Figure `r fig_schematic`. Schematic summarizing typical vertical gradients in (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) tree and ecosystem ecology within closed-canopy forests during daytime growing season conditions.** Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent. Patterns tend to be weaker, or sometimes reversed, in more open forests, when canopy trees are seasonally deciduous, or at nighttime. Abbreviations are as follows: $T_{air}$: air temperature, VPD: vapor pressure deficit;  $T_{leaf} - T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area; T sensitivity: temperature sensitivity; ET: evapotranspiration. *Isoprene emission has also been observed to peak in mid-canopy (Table `r table_leaf_traits`).](schematics/Schematic figure 1.png)

@@

Anatomical, structural and biochemical leaf traits vary vertically across forest strata [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, metabolism, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004]. 
Vertical variation in leaf structure and composition has long been recognized, with sun and shade leaf traits distinguished for over a century [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
More recent work has extended the focus from light alone to height in the vertical profile, as the latter shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
Overall, leaf biochemistry, anatomy and structure may be optimized to local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007]. 
However, we lack a a cohesive framework for integrating the many differences in leaf traits throughout the canopy.
Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004; @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020]. 
As canopy shade is known to alter these trait relationships [@osnasDivergentDriversLeaf2018], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.

@@

Leaf metabolism is strongly shaped by environmental drivers, $T_{leaf}$, and traits, all of which vary across vertical forest gradients (sections 1-3), Fig. `r fig_schematic`). 
However, as we detail below, few studies have evaluated how temperature responses of leaf metabolic rates are shaped by the varying environmental conditions across vertical forest gradients (Table `r table_leaf_metabolism`).

>>>>

#Word count
w_intro=369
w_biophys=1515
w_tleaf=749
w_traits= 2392
w_met=941
w_ecol=1353
w_vertgradients=w_biophys+w_tleaf+w_traits+w_met+w_ecol
w_imp_gc=1088
w_imp_modeling_scaling=1186
w_implications=w_imp_gc+w_imp_modeling_scaling
w_conclusions=248
w_total=w_intro+w_vertgradients+w_implications+w_conclusions

@@

![**Figure `r fig_schematic`. Schematic summarizing typical vertical gradients in (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf gas exchange, and (5) tree and ecosystem ecology within closed-canopy forests during daytime growing season conditions.** Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent. Patterns tend to be weaker, or sometimes reversed, in more open forests, when canopy trees are seasonally deciduous, or at nighttime. Abbreviations are as follows: $T_{air}$: air temperature, VPD: vapor pressure deficit;  $T_{leaf} - T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area; T sensitivity: temperature sensitivity; ET: evapotranspiration. *Isoprene emission has also been observed to peak in mid-canopy (Table `r table_leaf_traits`).](schematics/Schematic figure 1.png)

@@

Anatomical, structural and biochemical leaf traits vary vertically across forest strata [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, gas exchange, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004]. 
Vertical variation in leaf structure and composition has long been recognized, with sun and shade leaf traits distinguished for over a century [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
More recent work has extended the focus from light alone to height in the vertical profile, as the latter shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
Overall, leaf biochemistry, anatomy and structure may be optimized to local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007]. 
However, we lack a a cohesive framework for integrating the many differences in leaf traits throughout the canopy.
Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004; @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020]. 
As canopy shade is known to alter these trait relationships [@osnasDivergentDriversLeaf2018], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.

@@

Leaf gas exchange is strongly shaped by environmental drivers, $T_{leaf}$, and traits, all of which vary across vertical forest gradients (sections 1-3), Fig. `r fig_schematic`). 
However, as we detail below, few studies have evaluated how temperature responses of leaf metabolic rates are shaped by the varying environmental conditions across vertical forest gradients (Table `r table_leaf_metabolism`).
